---
title: "Cyclistic Bike-share Analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a case study completed as part of the Google Data Analytics Professional Certificate program. The goal here is to analyze the customer data of a bike-share company (Cyclistic) and assess the differences between casual riders (purchasers of single rides or day passes) and Cyclistic members (purchasers of annual memberships). The ultimate goal would be to create a digital media campaign aimed at converting casual riders in Cyclistic members. So let's get to it!

### Importing libraries and data
The data we'll be looking at is public bike-share made available by Motivate International Inc. at <https://divvy-tripdata.s3.amazonaws.com/index.html>, but for our purposes will be used to represent the fictional business Cyclistic. Since the data doesn't come from a third party source, we'll operate under the assumption that it's accurate. 
```{r warning=FALSE, message=FALSE}
# using the tidyverse collection of libraries to import and look through the most recent month's data
library(tidyverse)
library(ggmap)
trip_data <- read_csv("trip-data/202305-divvy-tripdata.csv")
```
### Processing the data
With the entire data set loaded, let's see what we have:
```{r}
head(trip_data)
glimpse(trip_data)
```
Finding some missing values, how many do we have in each column?
```{r}
# summing up NA values for every column
sapply(trip_data, function(x) sum(is.na(x)))
```
It looks like about 15% of both the start and end stations are missing values. Luckily we've also got geographic coordinates for these locations, of which only an insignificant number are missing values, so we can safely purge those entries. We'll get rid of the station name and id columns, as well as ride_id since this just appears to be a unique id for every entry.
```{r}
# selecting desired columns and filtering out rows with NA values
trip_data_clean <- trip_data %>%
  select(
    c(!ride_id & 
        !start_station_name &
        !start_station_id &
        !end_station_name &
        !end_station_id)
  ) %>%
  filter(c(!is.na(end_lat) & !is.na(end_lng)))
# rechecking NA values
sapply(trip_data_clean, function(x) sum(is.na(x)))
```
Perfect. Before moving on let's see what kind of values we have for rideable_type and member_casual.
```{r}
distinct(trip_data_clean, rideable_type)
distinct(trip_data_clean, member_casual)
```
Okay, it looks like we have 3 bike types and the 2 expected member levels; now we can infer some data from what we have, such as ride length and day of the week the ride was started.
```{r}
# adding length of ride as ride_minutes and weekday started as day_of_week
trip_data_rich <- trip_data_clean %>%
  mutate(
    ride_minutes = as.numeric(ended_at - started_at) / 60,
    day_of_week = weekdays(started_at)
  )
# taking a look at the new data set
glimpse(trip_data_rich)
summary(trip_data_rich)
```
Okay it looks like there are some negative ride lengths, so at least some of the ride times weren't accurately recorded. Given that the lowest duration is less than 2 minutes, it's most likely just a case of station clocks not being synced to the second, which isn't too much of a problem for our purposes, but we'll purge anything with a negative duration.
```{r}
trip_data_rich <- trip_data_rich %>%
  filter(ride_minutes >= 0)
```

### Analyzing the data
With our data in order, we can start to see how the different member types actually vary. First we'll see what rideable preferences each member type has.
```{r}
trip_data_rich %>%
  ggplot() +
    geom_bar(aes(member_casual, bg = rideable_type), position = "fill") +
    labs(
      title = "Rideable preference by member type",
      x = "Member type",
      y = "Preference ratio"
  )
```

It appears both groups have very similar bike preferences, with the exception Cyclistic members don't seem to use docked bikes at all (perhaps this isn't offered with the annual membership). Let's see how long both groups rides each rideable.
```{r}
trip_data_rich %>%
  group_by(rideable_type, member_casual) %>%
  reframe(
    ride_minutes = mean(ride_minutes)
  ) %>%
  ggplot() +
    geom_col(aes(ride_minutes, member_casual, bg = rideable_type), position = "dodge") +
    labs(
      title = "Average ride time by member type",
      x = "Duration in minutes",
      y = "Member Type"
    )
```

We can see that casual riders use rideables for longer than Cyclistic members, but also that docked bikes tend to be used for the longest rides. Now to see when they're riding.
```{r}
ordered_days <- factor(trip_data_rich$day_of_week, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
trip_data_rich %>%
  group_by(day_of_week, member_casual) %>%
  ggplot() +
    geom_bin_2d(aes(ordered_days, member_casual)) +
    labs(
      title = "Rideable usage per weekday",
      x = "Day of week",
      y = "Member type"
    ) +
    guides(x = guide_axis(angle = 45))
```

Here the difference between member types is seen most clearly. While both groups ride a similar amount on weekends, the Cyclistic members make much heavier use of rideables on weekdays, presumably as transportation to work or school, while at the same time casual riders are hardly using rideables at all. Finally, we'll take a look at where rides are starting and ending.
```{r message=FALSE, warning=FALSE}
map <- get_stamenmap(
  bbox = c(
    left = min(trip_data_rich$start_lng),
    bottom = min(trip_data_rich$start_lat),
    right = max(trip_data_rich$start_lng),
    top = max(trip_data_rich$start_lat)
  ),
  maptype = "terrain"
)
ggmap(map) + trip_data_rich %>%
  geom_point(mapping = aes(x = start_lng, y = start_lat, color = member_casual), position = "jitter", alpha = .5) +
  labs(
    title = "Starting locations",
    x = "Longitude",
    y = "Latitude"
  )
ggmap(map) + trip_data_rich %>%
  geom_point(mapping = aes(x = end_lng, y = end_lat, color = member_casual), position = "jitter", alpha = .5) +
  labs(
    title = "Ending locations",
    x = "Longitude",
    y = "Latitude"
  )
```

It looks like Cyclistic members tend to be much closer to the downtown area and nearby denser residential areas, while casual riders appear much more in the suburbs.

### Conclusion
The data here shows that the largest differences between member types is weekday usage and which part of the metro area they ride. My suggestion for a next step would be to investigate casual riders' need or interest in riding on weekdays, and perhaps focusing surveying and campaigning efforts in the suburban outer reaches of the area.